---
  title: "csDEX vignette"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{csDEX vignette}
%\VignetteEngine{knitr::rmarkdown}
\usepackage[utf8]{inputenc}
---
  
```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(dplyr)
```

One of the reasons that dplyr is fast is that it's very careful about when to make copies. This section describes how this works, and gives you some useful tools for understanding the memory usage of data frames in R.

The first tool we'll use is `dplyr::location()`. It tells us the memory location of three components of a data frame object:
  
* count (Data represented as features and their respective counts)
* PSI (Features are associaed to the percent spliced IN statistic)
* each attribute

There are two small datasets provided with the csDEX package, to be used for testing purposes only.

```{r}
  csDEXDataset(...)
```